WORKING TITLE: The adaptive value of green nest material across birds: A systematic review and meta-analysis

### 2025-07-26 manuscript_GNM.qmd
Temp commit back to desktop

Nest and nest-building are fundamental across a diverse range of animals, including arthropods (e.g., termites, spiders), mammals (e.g., platypus, rodents), and perhaps more famously, birds. Nests provide structural support for eggs, parents, and offspring, and protect them against predators and environmental pressures, and are thus key to reproductive success and survival [@hansell2000; @mainwaring2014; @deeming2023]. While most birds use organic plant materials such as twigs, branches, grass, and moss, some incorporate inorganic items, including mud, stones, feathers, snake skins and even anthropogenic materials like plastic and cigarette butts [@sergio2011; @suárez-rodríguez2013; @sheard2024]. These are considered non-structural additions and suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions. Indeed, nests can play a key role in sexual display, and features such as size have been shown to act as sexually selected signals of parental quality in several bird species [@moreno1994; @hoi1996], even interspecifically [@soler1995]. Nest-building behaviour is increasingly recognized as an expression of the extended phenotype (reviewed by @moreno2012), and phylogeny has been shown to be an important factor explaining both mean and within-species variation in nest morphology [@perez2023]. Yet, despite hundreds of scientific articles studying nests and nesting behaviour in birds being published each year, much remains unknown.

An intriguing nesting behaviour observed in some bird species is the addition of fresh, often aromatic, plant material to the nest. This green nest material (hereafter GNM), which often corresponds to leaves, sprigs, herbs and branches, is typically added from the start of nest building until egg laying, although some species like the blue tit (Cyanistes cyanus), keep adding it even after the nest is finished. Importantly, contrary to the structural nest materials (e.g., twigs, mud), GNM is selected from a small, non-random subset of local flora [@clark1985; @petit2002; @garrido-bautista2023], and often characterized by high levels of aromatic volatile compounds (commonly used by songbirds: @mennerat2009b) or secondary metabolites such as oils and resins (commonly used by raptors: @ontiveros2007). GNM is unlikely to be gathered opportunistically, and collecting these uncommon GNM, and replenishing it for some times weeks, requires extra time, and in some cases flying outside one’s territory [@mennerat2009]. Thus, the general costs associated with nest-building [@mainwaring2013], and the additional effort, time and energy devoted to add these non-structural materials suggest an important adaptive function of GNM.

@@

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; Clark and Mason 1988). Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results. In some cases, GNM was shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypothesis suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic (i.e., volatiles and secondary metabolites) properties have been also proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given that nest building is widespread across taxa —including but not limited to birds— and is increasingly recognized as an expression of the extended phenotype, a quantitative synthesis of this field is both timely and has potentially far-reaching implications.

@@

We conducted a systematic search on Web of Science Core Collection and Scopus on the 7th of September, 2022 (see update below). The keyword search query was designed as a three-block query including terms related to our treatment of interest ((“green\*” OR “herb\*” OR “aromatic\*”) AND “nest\*”), population of interest (“bird\*” OR “aves” OR “avian” OR “ornithol\*” OR “passerine\*” OR “passeriform\*” OR “songbird\*” OR list of all bird genera) and to limit the search results to experimental studies only (“experiment\*” OR “manipulat\*”). The complete search strings tailored to each specific platform are provided in @figsupp-string. Although our search string only included keywords in English, we planned to screen any study in English, Spanish, Portuguese and Hindi, but found none. We searched through titles, abstracts and keywords with no restrictions on publication date or study type. We validated our search using an initial library of 15 relevant articles that we had previously identified (@figsupp-own-library), ensuring all were retrieved. Additionally, we used the R package ‘litsearchr’ v1.0.0 [@grames2019], which did not yield any new relevant terms, confirm our queries. After removing `{r} nrow(full_reference_before_deduplication)-nrow(unique_reference_list)` duplicates using the R package ‘revtools’ v0.4.1 [@westgate2019], we identified `{r} nrow(unique_reference_list)` unique records.

@@

All titles and abstracts were equally allocated and screened independently by at least two screeners (SD, TR, JMGS) using the R package ‘revtools’ following a pre-designed decision tree with all inclusion/exclusion criteria (@figsupp-decision-tree). We included all studies on birds that experimentally manipulated the GNM (both aromatics or herbs). `{r} tools::toTitleCase(as.character(english::english(nrow(selected_abstract_screening))))` articles passed the title-and-abstract screening and were further evaluated during the full-text screening, where each article was independently screened by a single screener (SD, TR, JMGS), with another screener verifying the procedure. We used the same criteria as above (@figsupp-decision-tree) with an additional evaluation of whether the variable of interest in the article was in our list of fitness proxies of interest (@figsupp-fitness-proxies). `{r} paste0(toupper(substring(x <- as.character(english::english(nrow(selected_abstract_screening) - nrow(excluded_fulltext))), 1, 1)), substring(x, 2))` articles passed the full-text screening, and were, thus, eligible for our meta-analysis. All conflicts in screening decisions (1% for title-and-abstract screening, and none for full-text screening) were resolved through discussion.

We updated our systematic literature search by repeating our search on the 12th of August, 2024, and found `{r} nrow(unique_reference_list_repeat)` unique articles after removing `{r} nrow(full_reference_before_deduplication_repeat)-nrow(unique_reference_list_repeat)` duplicates. Only one observer (SD) performed the both screenings using the same decision trees as before (@figsupp-decision-tree), and found `{r} as.character(english::english(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T), 1, 1))` additional eligible article for our meta-analysis. Lastly, we identified `{r} as.character(english::english(nrow(selected_abstract_screening_repeat)-(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T)), 1, 1))` additional eligible articles recommended by two colleagues. In all, `{r} nrow(selected_abstract_screening)-nrow(excluded_fulltext)+nrow(selected_abstract_screening_repeat)` published articles were eligible for our meta-analysis, but we also obtained two additional unpublished studies (more details below and in @figsupp-prisma-flowchart).

@@

The data of each article were extracted evenly by one of three authors (SD, TR,  JMGS), with another screener verifying the procedure, and four authors (SD, TR, JMGS, AST) collectively discussing and revising unclear cases. We prioritised extracting primary data (i.e., measures of central tendency, variance and sample size) from text and tables, or, if that was not possible, from figures (e.g., boxplots) using the R package ‘metaDigitise’ v1.0.1 [@pick2018]. If primary data were not available, we extracted inferential statistics such as *t* and *F* values or descriptive statistics such as contingency tables (see below), that could be transformed into an effect size. Last, if the previous steps failed but the article had openly archived data, we calculated the means, standard deviation and sample sizes from them (n = `{r} length(raw_data)`).

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator “experimental design”. However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these `{r} nrow(lnRR_comp3)` and `{r} nrow(SMDH_comp3)` effect sizes, respectively, as non-aromatic GNM vs. no added GNM in our moderator “experimental design”. Importantly, our conclusions were robust to these effect sizes (\@Table SX), so we kept them in our dataset, and the moderator “experimental design” was used for subsequent meta-regressions. In studies where three groups were compared (n = `{r} nrow(multiple_comparision)`), we extracted all possible pairwise comparisons.

@@

Using the means, standard deviations and sample sizes for each control-treatment pairwise comparison, we calculated two effect sizes: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the escalc function (measure = ROM and SMDH respectively, vtype = "LS") in the R package ‘metafor’ v4.8-0 [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR, and treatment minus control for SMDH, such that positive values indicate a ‘fitness’ increase in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, or scab score, the effect size sign was inverted prior to analysis to ensure that a positive value consistently meant an increase in fitness regardless of fitness proxy (i.e., to ensure comparability across effect sizes). SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance; specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1. There were `{r} as.character(english::english(nrow(low_sample), 1, 1))` effect sizes from `{r} as.character(english::english(length(unique(low_sample$paper_ID)), 1, 1))`) study that did not fulfil that requirement, which we then excluded entirely from the dataset.

@@

When calculating each effect size’s associated sampling variance, we prioritised the number of nests in each group over the number of individuals as the sample size whenever possible to reduce the effect of genetic relatedness non-independence. In cases where the same control or treatment group was used across multiple comparisons, sample sizes were adjusted prior to calculating effect sizes by dividing them by the number of times each group was used in a comparison to account for shared group non-independence. For cases where the authors of an original article reported a statistically non-significant difference between two groups but without providing numerical values (k = `{r} nrow(no_ES)`, n = `{r} length(unique(no_ES$paper_ID))`), we assigned zero as the effect size for that comparison and imputed its sampling variance using the missing-case approach described in @nakagawa2022 for lnRR and using $\frac{n_1+n_2}{n_1×n_2}$ for SMD [@lajeunesse2013].

@@

We performed all analyses in R version 4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fitted using ‘metafor’ v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (ρ) between sampling variances from the same study [@noble2017] using the vcalc function from ‘metafor’ (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

@@

To test the relative importance of the two main adaptive hypotheses suggested for GNM, that is, the courtship and the parental care hypothesis, which combines the nest protection and drug hypotheses, we ran a multilevel meta-regression including the moderator ‘hypothesis’ (levels: courtship, parental care, both). For this and all meta-regressions, we estimated $R_{marginal}^2$,which corresponds to the percentage of heterogeneity explained by the moderators [@nakagawa2012]. Unless stated otherwise, all meta-regressions correspond to unimoderator meta-regressions.

@@

In addition, we performed X additional multilevel meta-regressions intended to explore the potential threats to the internal validity of each study on our conclusions (i.e., Risk of Bias or RoB) following recommendations in Culina et al. (2025). …

@@

We obtained `{r} nrow(dataset_analysis)` effect sizes from a total of `{r} length(unique(dataset_analysis$paper_ID))` studies conducted on `{r} length(unique(dataset_analysis$bird_species))` species across `{r} length(unique(dataset_analysis$population_ID))` geographical locations (@figmain-distribution). About `{r} round(sum(birds_info$perc[birds_info$perc>15]))`% of the effect sizes came from three bird species (`{r} birds_info$bird_species[birds_info$perc > 15][1]`, k = `{r} birds_info$n[birds_info$perc > 15][1]`; `{r} birds_info$bird_species[birds_info$perc > 15][2]`, k = `{r} birds_info$n[birds_info$perc > 15][2]`; `{r} birds_info$bird_species[birds_info$perc > 15][3]`, k = `{r} birds_info$n[birds_info$perc > 15][3]`). Our dataset included `{r} length(unique(unpublished_studies$paper_ID))` unpublished studies (k = `{r} nrow(unpublished_studies)`), with the remaining studies being published between `{r} min((dataset_analysis$year_publication),na.rm = T)` and `{r} max((dataset_analysis$year_publication),na.rm = T)`. The mean ±standard deviation number of effect sizes per study was `{r} sprintf("%.1f", mean(ES_per_study$n))` ± `{r} sprintf("%.1f",sd(ES_per_study$n))` (median: `{r} sprintf("%.1f", median(ES_per_study$n))` range: `{r} min(ES_per_study$n)`--`{r} max(ES_per_study$n)`) and the mean effective sample size (i.e., the number of nests/individuals per effect size after accounting for shared-group non-independence) was `{r} sprintf("%.1f",mean(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))` ±`{r} sprintf("%.1f", sd(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))` (median: `{r} sprintf("%.1f",median(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))`, range: `{r} min(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment)`--`{r} max(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment)`; `{r} nrow(dataset_analysis%>%filter(n_experiment<5|n_control<5))` data points had a sample size of less than 5 nests in each group).

@@

First, lnRR showed an average statistically non-significant increase in fitness outcomes of `{r} round((exp(as.double(coef(intercept_lnRR)))-1)*100,1)`% in nests with an experimental increase in GNM (@figmain-mean-effect A; lnRR = `{r} sprintf("%.3f", coef(intercept_lnRR))`, 95% CI: \[`{r} sprintf("%.3f", intercept_lnRR$ci.lb)` to `{r} sprintf("%.3f", intercept_lnRR$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_lnRR)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_lnRR)$pi.ub)`\], p-value = `{r} sprintf("%.3f", intercept_lnRR$pval)`, k = `{r} intercept_lnRR$k`, n = `{r} length(intercept_lnRR$s.levels.f[[1]])`) but heterogeneity among effect sizes was high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_lnRR$sigma2))` (Q = `{r} sprintf("%.1f",sum(intercept_lnRR$QE))`, p \< 0.001), the measures of relative (I^2^~total~ = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_lnRR)[1])`%) and standardised heterogeneity (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_lnRR)[1])` and M2~total =~ `{r} sprintf("%.2f", orchaRd::m2_ml(intercept_lnRR)[1])`) suggested high heterogeneity across effect sizes.

>>>>

Nest and nest-building are fundamental across a diverse range of animals, including arthropods (e.g., termites, spiders), mammals (e.g., platypus, rodents), and perhaps more famously, birds. Nests provide structural support for eggs, parents, and offspring, and protect them against predators and environmental pressures, and are thus key to reproductive success and survival [@hansell2000; @mainwaring2014; @deeming2023]. While most birds use organic plant materials such as twigs, branches, grass, and moss, some incorporate inorganic items, including mud, stones, feathers, snake skins and even anthropogenic materials like plastic and cigarette butts [@sergio2011; @suárez-rodríguez2013; @jagiello2023; @sheard2024]. These are considered non-structural additions and suggest that nests are more than just physical shelters; they can be complex tools shaped by both natural and sexual selection for multiple functions. Indeed, nests can play a key role in sexual display, and features such as size have been shown to act as sexually selected signals of parental quality in several bird species [@moreno1994; @hoi1996], even interspecifically [@soler1995]. Nest-building behaviour is increasingly recognized as an expression of the extended phenotype (reviewed by @moreno2012), and phylogeny has been shown to be an important factor explaining both mean and within-species variation in nest morphology [@perez2023]. Yet, despite hundreds of scientific articles studying nests and nesting behaviour in birds being published each year, much remains unknown.

An intriguing nesting behaviour observed in some bird species is the addition of fresh, often aromatic, plant material to the nest. This green nest material (hereafter GNM), which often corresponds to leaves, sprigs, herbs and branches, is typically added from the start of nest building until egg laying, although some species like the blue tit (Cyanistes cyanus), keep adding it even after the nest is finished. Importantly, contrary to the structural nest materials (e.g., twigs, mud), GNM is selected from a small, non-random subset of local flora [@clark1985; @petit2002; @garrido-bautista2023], and often characterized by high levels of aromatic volatile compounds (commonly used by songbirds: @mennerat2009b) or secondary metabolites such as oils and resins (commonly used by raptors: @ontiveros2007). GNM is unlikely to be gathered opportunistically, and collecting these uncommon GNM, and replenishing it for sometimes over weeks, requires extra time and, in some cases, flying outside one's territory [@mennerat2009]. Thus, the general costs associated with nest-building [@mainwaring2013], and the additional effort, time and energy devoted to add these non-structural materials suggest an important adaptive function of GNM.

@@

Second, the most widely tested hypothesis is the nest protection hypothesis, which proposes that the volatile compounds in GNM reduce the amount of parasites and pathogens in the nests, enhancing offspring survival. This hypothesis traces back to a comparative study in raptors (Falconiformes) that provided indirect evidence for it by comparing nest reuse, diet composition and GNM presence [@wimberger1984]. Soon after, a preference for GNM containing insect-repellent and antimicrobial volatiles was experimentally confirmed in European Starlings, therefore confirming two of the three key criteria for evaluating the nest protection hypothesis (i.e., selective plant use, bioactive chemical properties and demonstrable adaptive benefits; Clark and Mason 1988). Observational and experimental studies on raptors [@ontiveros2007; @dykstra2009] and passerines [@lafuma2001; @dawson2004; @mennerat2009b; @quiroga2012] have shown mixed results. In some cases, GNM was shown to improve offspring condition without reducing the amount of parasites in the nests [@gwinner2000], which led to the development of the drug hypothesis. This alternative hypothesis suggests that the volatiles present in GNM directly enhance offspring immune function and/or health independent of parasite or pathogen control [@gwinner2000; @mennerat2009a]. Thus, while both the nest protection and drug hypotheses suggest an adaptive benefit of GNM by enhancing offspring survival, the drug hypothesis emphasizes a pharmacological mechanism, whereas the nest protection hypothesis emphasizes parasite and pathogen repellency. Importantly, these two hypotheses are not mutually exclusive unless the improvement in offspring health can be proved to be unrelated to a reduction in parasite or pathogen load, something difficult to achieve.

Last, several additional adaptive hypotheses or functions of GNM that do not refer to its aromatic (i.e., volatiles and secondary metabolites) properties have also been proposed. These include benefits such as enhancing the crypticity of the nest [@skutch1957], providing insulation or sun protection for eggs and offspring [@mertens1977], or simply fulfilling a structural function [@lyons1987]. However, these alternative functional roles have received little support or attention. In all, despite decades of research, abundant observational and experimental studies, and two crucial narrative reviews [@dubiec2013; @scott-baumann2015], we still do not know the overall adaptive significance of GNM, its generality across bird species, and the relative importance of each of the suggested mechanisms. Given that nest building is widespread across taxa ---including but not limited to birds--- and is increasingly recognized as an expression of the extended phenotype, a quantitative synthesis of this field is both timely and has potentially far-reaching implications.

@@

We conducted a systematic search on Web of Science Core Collection and Scopus on the 7th of September, 2022 (see update below). The keyword search query was designed as a three-block query including terms related to our treatment of interest (("green\*" OR "herb\*" OR "aromatic\*") AND "nest\*"), population of interest ("bird\*" OR "aves" OR "avian" OR "ornithol\*" OR "passerine\*" OR "passeriform\*" OR "songbird\*" OR list of all bird genera) and to limit the search results to experimental studies only ("experiment\*" OR "manipulat\*"). The complete search strings tailored to each specific platform are provided in @figsupp-string. Although our search string only included keywords in English, we planned to screen any study in English, Spanish, Portuguese and Hindi, but found none. We searched through titles, abstracts and keywords with no restrictions on publication date or study type. We validated our search using an initial library of 15 relevant articles that we had previously identified (@figsupp-own-library), ensuring all were retrieved. Additionally, we used the R package 'litsearchr' v1.0.0 [@grames2019], which did not yield any new relevant terms, confirming our queries. After removing `{r} nrow(full_reference_before_deduplication)-nrow(unique_reference_list)` duplicates using the R package 'revtools' v0.4.1 [@westgate2019], we identified `{r} nrow(unique_reference_list)` unique records.

@@

All titles and abstracts were equally allocated and screened independently by at least two screeners (SD, TR, JMGS) using the R package 'revtools' following a pre-designed decision tree with all inclusion/exclusion criteria (@figsupp-decision-tree). We included all studies on birds that experimentally manipulated the GNM (both aromatics or herbs). `{r} tools::toTitleCase(as.character(english::english(nrow(selected_abstract_screening))))` articles passed the title-and-abstract screening and were further evaluated during the full-text screening, where each article was independently screened by a single screener (SD, TR, JMGS), with another screener verifying the procedure. We used the same criteria as above (@figsupp-decision-tree) with an additional evaluation of whether the variable of interest in the article was in our list of fitness proxies of interest (@figsupp-fitness-proxies). `{r} paste0(toupper(substring(x <- as.character(english::english(nrow(selected_abstract_screening) - nrow(excluded_fulltext))), 1, 1)), substring(x, 2))` articles passed the full-text screening, and were, thus, eligible for our meta-analysis. All conflicts in screening decisions (1% for title-and-abstract screening, and none for full-text screening) were resolved through discussion.

We updated our systematic literature search by repeating our search on the 12th of August, 2024, and found `{r} nrow(unique_reference_list_repeat)` unique articles after removing `{r} nrow(full_reference_before_deduplication_repeat)-nrow(unique_reference_list_repeat)` duplicates. Only one observer (SD) performed both screenings using the same decision trees as before (@figsupp-decision-tree), and found `{r} as.character(english::english(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T), 1, 1))` additional eligible article for our meta-analysis. Lastly, we identified `{r} as.character(english::english(nrow(selected_abstract_screening_repeat)-(sum(selected_abstract_screening_repeat$screened_abstracts=="selected", na.rm=T)), 1, 1))` additional eligible articles recommended by two colleagues. In all, `{r} nrow(selected_abstract_screening)-nrow(excluded_fulltext)+nrow(selected_abstract_screening_repeat)` published articles were eligible for our meta-analysis, but we also obtained two additional unpublished studies (more details below and in @figsupp-prisma-flowchart).

@@

The data of each article were extracted evenly by one of three authors (SD, TR,  JMGS), with another screener verifying the procedure, and four authors (SD, TR, JMGS, AST) collectively discussing and revising unclear cases. We prioritised extracting primary data (i.e., measures of central tendency, variance and sample size) from text and tables, or, if that was not possible, from figures (e.g., boxplots) using the R package 'metaDigitise' v1.0.1 [@pick2018]. If primary data were not available, we extracted inferential statistics such as t and F values or contingency tables to transform them into effect sizes (see below). Last, if the previous steps failed but the article had openly archived data, we calculated the means, standard deviation and sample sizes from each group ourselves (n = `{r} length(raw_data)`).

We extracted the means, variances and sample sizes for each control-treatment pairwise comparison. Depending on the experimental design of each study, controls included non-aromatic materials such as grass or moss, lesser amounts of aromatic material than the treatments, no additional material added, or removal of material added by birds; whereas treatments consisted of aromatic materials added by authors, birds, or both. Those comparisons were coded either as non-aromatic GNM vs. aromatic GNM, or no added GNM vs. aromatic GNM, in our moderator "experimental design". However, there was a third type of comparison used in some studies comparing nests with experimentally added non-aromatic GNM to nests without any added GNM or in which GNM added by the birds was removed by the authors. In this type of comparison, the non-aromatic GNM was considered the treatment, and we coded these `{r} nrow(lnRR_comp3)` and `{r} nrow(SMDH_comp3)` effect sizes, respectively, as non-aromatic GNM vs. no added GNM in our moderator "experimental design". Importantly, our conclusions were robust to these effect sizes (\@Table SX), so we kept them in our dataset, and the moderator "experimental design" was used for subsequent meta-regressions. In studies where three groups were compared (n = `{r} nrow(multiple_comparision)`), we extracted all possible pairwise comparisons.

@@

Using the means, standard deviations and sample sizes for each control-treatment pairwise comparison, we calculated two effect sizes: a log response ratio (lnRR) and a standardised mean difference with heteroskedasticity correction (SMDH), using the escalc function (measure = ROM and SMDH respectively, vtype = "LS") in the R package 'metafor' v4.8-0 [@metafor]. We specified the treatment group as the numerator and the control group as the denominator for lnRR, and treatment minus control for SMDH, such that positive values indicate a 'fitness' increase in the treatment group. In cases such as measures of parasite load or prevalence, laying or hatching dates, courtship time, or scab score, the effect size sign was inverted prior to analysis to ensure that a positive value consistently meant an increase in fitness regardless of fitness proxy (i.e., to ensure comparability across effect sizes). SMDH could only be calculated when there were enough data points in both treatment and control groups to estimate the pooled variance; specifically, when the degrees of freedom (i.e., n1+n2−2) was greater than 1. There were `{r} as.character(english::english(nrow(low_sample), 1, 1))` effect sizes from `{r} as.character(english::english(length(unique(low_sample$paper_ID)), 1, 1))`) study that did not fulfil that requirement, which we then excluded entirely from the dataset.

@@

When calculating each effect size's associated sampling variance, we prioritised the number of nests in each group over the number of individuals as the sample size whenever possible to reduce the effect of genetic relatedness non-independence. In cases where the same control or treatment group was used across multiple comparisons, sample sizes were adjusted prior to calculating effect sizes by dividing them by the number of times each group was used in a comparison to account for shared group non-independence. For cases where the authors of an original article reported a statistically non-significant difference between two groups but without providing numerical values (k = `{r} nrow(no_ES)`, n = `{r} length(unique(no_ES$paper_ID))`), we assigned zero as the effect size for that comparison and imputed its sampling variance using the missing-case approach described in @nakagawa2022 for lnRR and using $\frac{n_1+n_2}{n_1×n_2}$ for SMD [@lajeunesse2013].

@@

We performed all analyses in R version 4.5.1 [@R-version]. All multilevel meta-analytical models and meta-regressions were fitted using 'metafor' v.4.8-0 [@metafor]. Sampling variances were modelled as variance-covariance matrices assuming a 0.5 correlation (ρ) between sampling variances from the same study [@noble2017] using the vcalc function from 'metafor' (See Supplementary Appendix @figsupp-correlation for sensitivity analyses using different values of *ρ*).

@@

To test the relative importance of the two main adaptive hypotheses suggested for GNM, that is, the courtship and the parental care hypothesis, which combines the nest protection and drug hypotheses, we ran a multilevel meta-regression including the moderator 'hypothesis' (levels: courtship, parental care, both). For this and all meta-regressions, we estimated $R_{marginal}^2$,which corresponds to the percentage of heterogeneity explained by the moderators [@nakagawa2012]. Unless stated otherwise, all meta-regressions correspond to unimoderator meta-regressions.

@@

In addition, we performed X additional multilevel meta-regressions intended to explore the potential threats to the internal validity of each study on our conclusions (i.e., Risk of Bias or RoB) following recommendations in Culina et al. (2025). ...

@@

We obtained `{r} nrow(dataset_analysis)` effect sizes from a total of `{r} length(unique(dataset_analysis$paper_ID))` studies conducted on `{r} length(unique(dataset_analysis$bird_species))` species across `{r} length(unique(dataset_analysis$population_ID))` geographical locations (@figmain-distribution). About `{r} round(sum(birds_info$perc[birds_info$perc>15]))`% of the effect sizes came from three bird species (`{r} birds_info$bird_species[birds_info$perc > 15][1]`, k = `{r} birds_info$n[birds_info$perc > 15][1]`; `{r} birds_info$bird_species[birds_info$perc > 15][2]`, k = `{r} birds_info$n[birds_info$perc > 15][2]`; `{r} birds_info$bird_species[birds_info$perc > 15][3]`, k = `{r} birds_info$n[birds_info$perc > 15][3]`). Our dataset included `{r} length(unique(unpublished_studies$paper_ID))` unpublished studies (k = `{r} nrow(unpublished_studies)`), with the remaining studies being published between `{r} min((dataset_analysis$year_publication),na.rm = T)` and `{r} max((dataset_analysis$year_publication),na.rm = T)`. The mean ±standard deviation number of effect sizes per study was `{r} sprintf("%.1f", mean(ES_per_study$n))` ± `{r} sprintf("%.1f",sd(ES_per_study$n))` (median: `{r} sprintf("%.1f", median(ES_per_study$n))` range: `{r} min(ES_per_study$n)`--`{r} max(ES_per_study$n)`) and the mean effective sample size (i.e., the number of nests/individuals per effect size after accounting for shared-group non-independence) was `{r} sprintf("%.1f",mean(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))` ±`{r} sprintf("%.1f", sd(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))` (median: `{r} sprintf("%.1f",median(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))`, range: `{r} floor(min(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))`--`{r} round(max(dataset_analysis$effective_n_control+dataset_analysis$effective_n_experiment))`; `{r} nrow(dataset_analysis%>%filter(n_experiment<5|n_control<5))` data points had a sample size of less than 5 nests in each group).

@@

First, lnRR showed an average statistically non-significant increase in fitness outcomes of `{r} round((exp(as.double(coef(intercept_lnRR)))-1)*100,1)`% in nests with an experimental increase in GNM (@figmain-mean-effect A; lnRR = `{r} sprintf("%.3f", coef(intercept_lnRR))`, 95% CI: \[`{r} sprintf("%.3f", intercept_lnRR$ci.lb)` to `{r} sprintf("%.3f", intercept_lnRR$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_lnRR)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_lnRR)$pi.ub)`\], p-value = `{r} sprintf("%.3f", intercept_lnRR$pval)`, k = `{r} intercept_lnRR$k`, n = `{r} length(intercept_lnRR$s.levels.f[[1]])`) but heterogeneity among effect sizes was high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_lnRR$sigma2))` (*Q* = `{r} sprintf("%.1f",sum(intercept_lnRR$QE))`, p \< 0.001), the measures of relative (*I^2^~total~* = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_lnRR)[1])`%) and standardised heterogeneity (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_lnRR)[1])` and M2~total =~ `{r} sprintf("%.2f", orchaRd::m2_ml(intercept_lnRR)[1])`) suggested high heterogeneity across effect sizes.

Second, SMD(H) showed an average statistically significant increase in fitness outcomes in nests with an experimental increase of GNM (@figmain-mean-effect B; SMD(H) = `{r} sprintf("%.3f", coef(intercept_SMDH))`, 95% CI: \[`{r} sprintf("%.3f",(intercept_SMDH)$ci.lb)` to `{r} sprintf("%.3f",(intercept_SMDH)$ci.ub)`\], 95% PI: \[`{r} sprintf("%.3f", predict(intercept_SMDH)$pi.lb)` to `{r} sprintf("%.3f", predict(intercept_SMDH)$pi.ub)`\], p-value = `{r} sprintf("%.3f",intercept_SMDH$pval)`, k = `{r} intercept_SMDH$k`, n = `{r} length(intercept_SMDH$s.levels.f[[1]])`), but heterogeneity among effect sizes was moderate to high. Total raw heterogeneity (𝜎^2^) was `{r} sprintf("%.3f",sum(intercept_SMDH$sigma2))` (*Q* = `{r} sprintf("%.1f",sum(intercept_SMDH$QE))`, p \< 0.001), relative heterogeneity was smaller compared to that of lnRR (*I^2^~total~* = `{r} sprintf("%.1f",orchaRd::i2_ml(intercept_SMDH)[1])`%) and the standardised heterogeneity metrics (CVH2~total~ = `{r} sprintf("%.2f", orchaRd::cvh2_ml(intercept_SMDH)[1])` and M2~total =~ `{r} sprintf("%.2f",orchaRd::m2_ml(intercept_SMDH)[1])`) lied between the 50th and 75th percentile of empirically derived estimates from meta-analyses in ecology and evolution [@yang2024a]. For lnRR, all three heterogeneity metrics exceeded the 75th percentile benchmark (see [@supp-tab-heterogeneity] for detailed estimates of heterogeneity).

These overall results, and their conclusions were robust to several sensitivity analyses. First, we fitted a pre-registered bivariate multilevel meta-analytic model including both effect sizes as response variables (lnRR = `{r} sprintf("%.3f",(coef(bivariate_model)[1]))`, p-value = `{r} sprintf("%.3f",(bivariate_model$pval[1]))` and SMD(H) = `{r} sprintf("%.3f",(coef(bivariate_model)[2]))` , p-value = `{r} sprintf("%.3f",(bivariate_model$pval[2]))`). In addition, we ran several sensitivity analyses to test the robustness of the overall results to: (i) the inclusion of traits for which their relationship with fitness was debated in the literature (non-pre-registered), (ii) the removal effect sizes that were originally reported as simply statistically non-significant (non-pre-registered), (iii) both (i) and (ii) (non-pre-registered), and, (iv) the removal of effect sizes that were calculated by transforming inferential statistics. The number of effect sizes included in those sensitivity analyses ranged from 182 to 274 effect sizes, and the conclusions remained largely the same [@tabsupp-SA-intercept].
